A Bayesian networks approach to infer social changes in burials in northeastern Taiwan during the European colonization period

### 2021-06-21 analysis/paper/paper.Rmd
edits to the text after peer review and revisions

<!--Our burial network and research implication-->
Burial studies are usually based on characterizing each physical trace or burial variable, such as biological records, grave forms and goods, or ritual behaviors, and finally combining and comparing those individual observations to infer the organization of past societies [@Byrd1995; @Seikel2011]. Recent studies on burials using network analysis demonstrates it is a promising tool to test expectations derived from hypotheses [@Sosna2013]. Here we use network modeling to explore similarities and differences between burials by identifying underlying variable structuring burial relations. We treat burials as a complex network based on the assumption that a network is a patterned aggregation that includes individual elements (i.e. individual burials), pairwise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing social patterns represented in the data [@Brandes2013]. Our network indicates status similarity between individuals, where a tie means two individuals sharing a similar social status or a similar social environment that higher status individuals will be given similar burial goods [eg. @Collar2015, pp. 4; eg. @Sosna2013]. This is because burial goods, especially high value ones, normally convey cultural symbol of personal wealth or social status from which we can infer social differentiation [@Gamble2002; @Janes2013]. Our case study illustrates how a Bayesian approach to network modeling provides new insights into the formation of archaeological networks and allows testing of anthropological models to better understand dynamic social processes.

@@

<!--theoretical framework: corporate/network strategies-->
The involvement in long-distance exchange is often associated with an increased social inequality where ambitious individuals could gain power by building personal influence through the acquisition and distribution of foreign goods, leading stoical change from corporate based to network based societies [@Blanton1996; @Feinman2000; @Klehm2017]. Corporate-network societies represent two distinct political-economic strategies for obtaining power, viewed as a horizontal axis crosscutting traditional vertical hierarchy [@Drennan2010; @Feinman2000]. The features of network strategies include the accumulation of prestige goods, wealth differentiation, and trade monopolization through individual networks. These are contrasted with corporate strategies that stress communal ritual elements, shared power, and where wealth differentiation, if any, would be associated with corporate groups, such as age-based or kin-based, rather than individuals [@Siegel1999]. In Asia, network-based societies are identified, especially for places involved in long-distance trade [@Carter2015; @Ueda2016; @Liu2006].

<!--archaeological context of Kiwulan-->
The observed uneven distribution of burial goods at Kiwulan may be explained as a result of unequal access to trade goods when Indigenous societies became more involved in the complex trade network stimulated by European colonial activities in northern Taiwan. Foreign trade goods were believed to be used as prestige goods based on archaeological contexts at Kiwulan [@Hsieh2009; @Wang2011]. There are very few trade goods found in middens but more in residential area and burials, along with a symbol of personal influence described in the Spanish historical records [@LiandWu2006]. In addition, the study of spatial distribution of trade ornaments in Kiwulan residential area demonstrates a more clustered pattern after the European presence, indicative of intentional accumulation of trade goods [@LWandBM2020ornament]. The European contact and associated trade activities could be the stimulus of competition for foreign resources among aggrandizers that led to social changes.

@@

<!--assigning burials based on our chronology-->
To compare burial networks, we assigned burials to the pre-European period (n = `r pre_E`), European and post-European period (n = `r post_E`). Our assignments are based on an established fine-grained chronology that was reexamined and cross-validated by diagnostic materials, stratigraphic data, depth, and radiocarbon ages [for more details see @LWandBM2020ornament; @LWandBM2020pot]. We excluded burials from the Chinese phase (n = `r chi`) due to the smaller sample size. We also excluded `r disturb` burials that were heavily disturbed by modern construction which prevented accurate determination of their chronology. Differences in network size were evaluated using vertex bootstrap in a later section. By comparing networks from the pre-European period and the post-European period we can examine the effects of foreign contact on community relationships at Kiwulan.

@@

<!--the context of burial networks-->
We built networks where burials (nodes in the network) are linked when they have the same trade goods in common. This is based on the assumptions that social status and associated relations can be represented by sharing similar goods and burials can reflect social structures [@Binford1971; @Saxe1970; cf. @Coward2013, pp. 252]. Thus, individuals with the same prestige goods reflect similar access to trade, exchange, and gifting networks, indicative of the similarity among Kiwulan residents and social inequality. The trade goods we identified include gold-foil beads, carnelian beads, glass beads, Chinese porcelains, stonewares, gold foils, fish-shaped ornaments, and imported coins, which were also items less affected by taphonomic processes because of durable materials. Those goods are considered as prestige goods due to their rarity in burial contexts and descriptions in historical records [@Chen2007; @Cheng2008; @Hsieh2009; @Hsieh2012; @Wang2011]. European historical accounts mentioned that those items were treated as prestige goods in Indigenous culture [@Borao2009; @LiandWu2006]. For example, a Spanish priest described a Spanish soldier exchanging carnelian beads for natural resources with local Indigenous people, because of the beads’ high value in Indigenous culture [@LiandWu2006]. In addition, Spanish visitors observed that Indigenous people possessing more imported goods were recognized as having higher status in their community [@LiandWu2006].

@@

<!-- add text about prior setting and anthropological assumptions of our prior values --> 
Priors are the distinctive aspect of Bayesian methods to estimate population by combining one's beliefs with data [@Gill2014]. Normal distribution N($\mu$, $\sigma^2$), where $\mu$ is mean and $\sigma^2$ is standard deviation, is typical for prior specification to model network parameters because it is sensible and convenient, allowing adjustments for a wide range of prior information [@Caimo2017; @Normand1992]. The larger the mean of the prior distribution, the more positive effect of that parameter on network formation. For standard deviation, if it is larger, it presents a relatively vague prior enabling comparison between variables that we do not have specific assumptions. Since low density is commonly found in the real world [@Caimo2017], we specified the prior of the edge density parameter to low for both network models, following a normal distribution with mean at `r priors$mean_same[1]`, and standard deviations at `r priors$sigma_same[1]` (i.e (N(`r priors$mean_same[1]`, `r priors$sigma_same[1]`))).

To evaluate our anthropological model of increased social inequality over time to indicate a shift from a corporate to a network society [@Drennan2010; @Feinman2000], we specify priors that are meaningful for social change, especially for the variables of transitivity and centralization. We set the priors to higher transitivity (N(`r priors$pre_prior_mean[6]`, `r priors$pre_prior_sigma[6]`)), lower centralization (N(`r priors$pre_prior_mean[7]`, `r priors$pre_prior_sigma[7]`)), and higher covariate effect of ritual activity (N(`r priors$pre_prior_mean[4]`, `r priors$pre_prior_sigma[4]`)) for the network before European contact to indicate less social differentiation and emphasize the ritual element shared in corporate groups. Conversely, we set the priors for the network after European arrival to lower transitivity (N(`r priors$post_prior_mean[6]`, `r priors$post_prior_sigma[6]`)), higher centralization (N(`r priors$post_prior_mean[7]`, `r priors$post_prior_sigma[7]`)), and a higher covariate effect for burial values (N(`r priors$post_prior_mean[5]`, `r priors$post_prior_sigma[5]`)) to model an increased social differentiation after European contact. For covariates about biological features, such as age, sex, we specified a vague prior (N(`r priors$mean_same[2]`, `r priors$sigma_same[2]`)) for both models to explore their effects. Similarly, we used the same prior (N(`r priors$mean_same[8]`, `r priors$sigma_same[8]`)) for physical distance between burials, to explore whether there is kinship-based proximity, e.g. stronger correlations for shorter distances. We then examine the fit of the observed data to the two models we specified to evaluate our hypotheses.

@@

The entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in the Supplementary Online Materials at http://doi.org/10.17605/OSF.IO/XGA6N to enable re-use of materials and improve reproducibility and transparency [@Marwick2017]. Also in this version-controlled compendium [@Marwick2018] are the raw data for all the visualizations and tests reported here and additional details of definition of our approach to tie formation, model fitting, assessment of MCMC output and goodness-of-fit diagnostics for our models. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.

@@

<!--posterior stats for pre-E-->
We examine the estimates from the posterior distributions to compare their differences in structure of the simulated networks (Table \@ref(tab:table-posterior-stats); Figure \@ref(fig:figure-posterior-distribution)). For nodal covariates in the pre-European model, the values of ritual element represented by pots (ritual-homophily) are mostly greater than zero that suggests a positive effect on the formation of relations between burials, while the values of wealth rank represented by burial values (wealth-homophily) are mostly negative that indicates it has less influence on the tie formation. This demonstrates that burials with ritual pottery tend to form relations, but burials in the same wealth level tend not to. For other covariates with values mostly around zero, such as age (age-homophily) and sex (sex-homophily), they do not show any clear tendency. For the physical distance variable (dyadic covariate), it shows that most values are positive, indicating that physical proximity between burials could reflect similarity in burial treatment. For the endogenous network effects, transitivity presents a significant positive effect, while centralization demonstrates a negative effect. The high positive value for transitivity suggests a tendency of burials with similar burial goods to be clustered as connected communities, indicative of the presence of multiple corporate groups sharing burial goods in common. In contrast, the strong negative centralization shows there is a tendency toward decentralization that reflects most burials having a similar number of ties without any prominent burials. This might imply that individuals have equal access to trade goods in terms of the flow of goods.

@@

library(flextable)
# add column labels
ft <- flextable(posterior_esti_two_phases) %>% 
  colformat_double(digits = 2) %>% 
  set_header_labels(mean_pre = "pre-European", median_pre = "pre-European", 
                    `2.5%_pre` = "pre-European", `97.5%_pre` = "pre-European",
                    mean_post = "post-European", median_post = "post-European", 
                    `2.5%_post` = "post-European", `97.5%_post` = "post-European") %>%
  merge_at(i = 1, j = 2:5, part = "header") %>% 
  merge_at(i = 1, j = 6:9, part = "header") %>% 
  add_header_row(values = c("", "mean", "median", "2.5%", "97.5%",
                            "mean", "median", "2.5%", "97.5%"), top = FALSE ) %>% 
  align(j = 2:9, align = "right", part = "header") %>% 
  align(i = 1, align = "center", part = "header") %>% 
  width(j = ~ parameter, width = 1.36) %>% 
  width(j = 2:9, width = 0.67) %>% 
  set_caption("Estimated posterior means, medians, and 95% confidence intervals for each network parameter of two models. Confidence intervals that do not include zero indicate significant effects of parameters (Caimo, 2017). Some posterior means close to posterior medians that indicate symmetric distributions.")
ft
```

@@

<!--posterior stats for for post-E and their comparison-->
For the post-European network model, two nodal covariates, wealth rank and age, both show tendencies towards positive values. This indicates the burials in the same wealth level and age groups tend to form relations. However, there is no clear pattern for ritual and sex variables since most values around zero. Also, there are no effects for the variable of physical distance. Similar to the pre-European network, transitivity demonstrates a significant positive effect, but weaker than the effect of the pre-European network. In contrast, centralization has a significantly higher positive effect than the effect of the pre-European network. This means there is a tendency toward centralization, reflecting a limited number of burials having many more ties than others. This implies that the presence of better access to trade goods and the behavior of wealth accumulation and display in burial events. In general, the post-European network model has a smaller transitivity effect and higher centralization effect than the pre-European network model. This may suggest a reduced tendency toward clustering but high tendency toward centralization after the European presence.

@@

<!-- comparison between two networks using bootstrap --> 
One key difference between the pre-European and post-European networks is their size, with `r pre_E` burials compared to `r post_E` burials. To understand the robustness of comparison between two networks, we used a vertex bootstrap technique to cross-validate the results of our Bayesian ERGMs. The vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. node) to quantify standard errors and estimate sampling variability in the network statistics of interest [@Chen2019; @Snijders1999; @Rroberts2021]. This enables the evaluation of uncertainty for networks and tests the difference between multiple networks of different sizes by examining their confidence intervals for the network population. We used the vertex bootstrap to compute endogenous network statistics, including density, centralization, and transitivity for our two networks. We explored the sample size effect by removing nodes at certain percentages (5-40%) for both networks and comparing their confidence intervals. The results demonstrate a consistent difference between the two networks up to 35% node removals for the centralization variable as shown in Figure \@ref(fig:figure-bootstrap-CI). This shows consistency with our earlier finding of negative centralization in the pre-European period and positive centralization in the post-European period, over several conditions of node removal. Similarly consistent with our earlier result, the vertex bootstrap shows no significant difference on density and transitivity. This indicates that the network variables, including density, centralization, and transitivity, are robust to the different network size of our sample.

@@

<!--archaeological implication based on the results-->
Network covariates give insights into the socio-cultural factors that were associated with the acquisition and distribution of foreign prestige goods. The general pattern of posterior distributions indicates that the main factor contributing to burial networks for the pre-European network was ritual, and for the post-European network was wealth. We interpret these results as illustrating different social and economic mechanisms influencing relations between burials for the two periods. Before European arrival, ritual behaviors could be important factors that structured the interconnections of prestige goods in the burial contexts. This might imply that ritual as a status indicator at the time before many foreign goods were introduced to Indigenous societies. After the European presence was established, status indicators shifted from ritual to wealth differences, corresponding with the characteristics of corporate/network strategies. In addition, geographical proximity of burials, implying kinship association, has some effect on sharing similar prestige items between burials in the pre-European period, but not in the post-European period. This may suggest that the use of prestige items is more individual-oriented instead of kinship-based after European arrival. Demographic variable, age, also indicates that burials in the same age may tend to have similar goods after European arrival. However, another demographic varialbe, sex, do not have any effects on network formation for both networks. We should also note that their effects may be impacted by the unavailability of age (`r age_miss` out of `r pre_and_post` burials) and sex (`r sex_miss` out of `r pre_and_post` burials) values in our sample.

@@

<!-- trade dynamic during this period and future work for comparison -->
We recognize that trade dynamics between different Indigenous groups and other foreign presence, Han Chinese and Japanese, might have contributed to social changes at Kiwulan as well. However, the current changes observed from burial networks and spatial distribution of trade items in the residential area at Kiwulan demonstrate consistent changes before and after the European arrival. The frequent trade activities stimulated by the presence of Europeans led to many trade items, including Anping jars, stonewares, Chinese porcelains, and beads found at archaeological sites with Dutch occupation at different parts of Taiwan, such as Fort Zeelandia. As a large village with continuous occupation since the 14th century in northeastern Taiwan, Kiwulan provides a special example to explore the reaction of Indigenous people living on the periphery to European colonial presence at a settlement scale. However, we still need more evidence from other sites in northeastern Taiwan to dive deeper into pericolonial effects, and compare evidence between sites on the periphery and colonial centers, such as Heping Dao and Tamsui. Unfortunately, Indigenous sites related to European contact with intensive excavation and sufficient evidence are rare to offer suitable comparison with Kiwulan. Future work is still needed.

>>>>

<!--Our burial network and research implication-->
Burial studies are usually based on characterizing each physical trace or burial variable, such as biological records, grave forms and goods, or ritual behaviors, and finally combining and comparing those individual observations to infer the organization of past societies [@Byrd1995; @Seikel2011]. Recent studies on burials using network analysis demonstrates it is a promising tool to test expectations derived from hypotheses [@Sosna2013]. Here we use network modeling to explore similarities and differences between burials by identifying underlying variable structuring burial relations. We treat burials as a complex network based on the assumption that a network is a patterned aggregation that includes individual elements (i.e. individual burials), pairwise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing social patterns represented in the data [@Brandes2013]. Our network indicates status similarity between individuals, where a tie means two individuals sharing a similar social status or a similar social environment that higher status individuals will be given similar burial goods [eg. @Collar2015, pp. 4; eg. @Sosna2013]. In this case, network ties do not indicate exchange or transmission but rather similarities. This is because burial goods, especially high value ones, are normally cultural symbols of personal wealth or social status from which we can infer social differentiation [@Gamble2002; @Janes2013]. Our case study illustrates how a Bayesian approach to network modeling provides new insights into the formation of archaeological networks and allows testing of anthropological models to better understand dynamic social processes.

@@

<!--theoretical framework: corporate/network strategies-->
The involvement in long-distance exchange is often associated with an increased social inequality where ambitious individuals could gain power by building personal influence through the acquisition and distribution of foreign goods, leading to social change from corporate to network societies [@Blanton1996; @Feinman2000; @Klehm2017]. Corporate-network societies represent two distinct political-economic strategies for obtaining power, viewed as a horizontal axis crosscutting traditional vertical hierarchy [@Drennan2010; @Feinman2000]. The features of network strategies include the accumulation of prestige goods, wealth differentiation, and trade monopolization through individual networks. These are contrasted with corporate strategies that stress communal ritual elements, shared power, and where wealth differentiation, if any, would be associated with corporate groups, such as age-based or kin-based, rather than individuals [@Siegel1999]. In Asia, network-based societies are frequently identified, especially for places involved in long-distance trade [@Carter2015; @Ueda2016; @Liu2006].

<!--archaeological context of Kiwulan-->
The observed uneven distribution of burial goods at Kiwulan may be explained as a result of unequal access to trade goods when Indigenous societies became more involved in the complex trade networks stimulated by European colonial activities in northern Taiwan. Foreign trade goods were believed to be used as prestige goods based on archaeological contexts at Kiwulan [@Hsieh2009; @Wang2011]. There are very few trade goods found in middens but more in the residential area and burials. According to Spanish historical records they are a symbol of personal influence [@LiandWu2006]. In addition, the study of spatial distribution of trade ornaments in Kiwulan residential area demonstrates a more clustered pattern after the European presence, indicative of intentional accumulation of trade goods [@LWandBM2020ornament]. The European contact and associated trade activities could be the stimulus of competition for foreign resources among aggrandizers that led to social changes.

@@

<!--assigning burials based on our chronology-->
To compare burial networks, we assigned burials to the pre-European period (n = `r pre_E`), European and post-European period (n = `r post_E`). Our assignments are based on an established fine-grained chronology that was reexamined and cross-validated by diagnostic materials, stratigraphic data, depth, and radiocarbon ages [for more details see @LWandBM2020ornament; @LWandBM2020pot]. We excluded burials from the Chinese phase (n = `r chi`) due to the smaller sample size. For the two periods studied here, we evaluate differences in network size were using a vertex bootstrap method. We also excluded `r disturb` burials that were heavily disturbed by modern construction which prevented accurate determination of their chronology. Most burials have clear stratigraphic contexts so modern or ancient disturbance is readily visible (i.e. if a burial has been cut into by later burials, very few cases of this have been documented at Kiwulan).  By comparing networks from the pre-European period and the post-European period we can examine the effects of foreign contact on community relationships at Kiwulan.

@@

<!--the context of burial networks-->
We built networks where burials (nodes in the network) are linked when they have the same trade goods in common. This is based on the assumptions that social status and associated relations can be represented by sharing similar goods and burials can reflect social structures [@Binford1971; @Saxe1970; cf. @Coward2013, pp. 252]. Thus, individuals with the same prestige goods reflect similar access to trade, exchange, and gifting networks, indicative of the similarity among Kiwulan residents and social inequality. The trade goods we identified include gold-foil beads, carnelian beads, glass beads, Chinese porcelains, stonewares, gold foils, fish-shaped ornaments, and imported coins, which were also items less affected by taphonomic processes because of their durable materials. Those goods are considered as prestige goods due to their rarity in burial contexts and descriptions in historical records [@Chen2007; @Cheng2008; @Hsieh2009; @Hsieh2012; @Wang2011]. European historical accounts mentioned that those items were treated as prestige goods in Indigenous culture [@Borao2009; @LiandWu2006]. For example, a Spanish priest described a Spanish soldier exchanging carnelian beads for natural resources with local Indigenous people, because of the beads’ high value in Indigenous culture [@LiandWu2006]. In addition, Spanish visitors observed that Indigenous people possessing more imported goods were recognized as having higher status in their community [@LiandWu2006].

@@

<!-- add text about prior setting and anthropological assumptions of our prior values --> 
Priors are a distinctive aspect of Bayesian methods that enable us combining our beliefs based on prior research with the newly observed data [@Gill2014]. The normal distribution N($\mu$, $\sigma^2$), where $\mu$ is the mean and $\sigma^2$ is the standard deviation, is typically used for priors when modelling network parameters because it is sensible and convenient, allowing adjustments for a wide range of prior information [@Caimo2017; @Normand1992]. The larger the mean of the prior distribution, the more positive effect of that parameter on network formation. A larger standard deviation makes the prior more vague, enabling comparison between variables for which we do not have specific assumptions. Since low network densities are commonly found in the real world [@Caimo2017], we specified the prior of the edge density parameter to low for both network models, following a normal distribution with mean at `r priors$mean_same[1]`, and standard deviations at `r priors$sigma_same[1]` (i.e (N(`r priors$mean_same[1]`, `r priors$sigma_same[1]`))).

To evaluate our anthropological model of increased social inequality, specifically a shift from a corporate to network society [@Drennan2010; @Feinman2000], we specified priors that are meaningful for social inequality, especially for the variables of transitivity and centralization. For the network before European contact, we set the priors to model less social differentiation and emphasize the ritual element shared in corporate groups: higher transitivity (N(`r priors$pre_prior_mean[6]`, `r priors$pre_prior_sigma[6]`)), lower centralization (N(`r priors$pre_prior_mean[7]`, `r priors$pre_prior_sigma[7]`)), and a higher covariate effect of ritual activity (N(`r priors$pre_prior_mean[4]`, `r priors$pre_prior_sigma[4]`)). Conversely, to model an increased social differentiation after European contact, we set the priors for the network after European arrival to lower transitivity (N(`r priors$post_prior_mean[6]`, `r priors$post_prior_sigma[6]`)), higher centralization (N(`r priors$post_prior_mean[7]`, `r priors$post_prior_sigma[7]`)), and a higher covariate effect for burial values (N(`r priors$post_prior_mean[5]`, `r priors$post_prior_sigma[5]`)).  For covariates about biological features, such as age, sex, we specified a vague prior (N(`r priors$mean_same[2]`, `r priors$sigma_same[2]`)) for both models to explore their effects. Similarly, we used the same vague prior (N(`r priors$mean_same[8]`, `r priors$sigma_same[8]`)) for physical distance between burials, to explore whether there is kinship-based proximity, e.g. stronger correlations for shorter distances. We then examine the fit of the observed data to the two models we specified to evaluate our hypotheses.

@@

The entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in the Supplementary Online Materials at http://doi.org/10.17605/OSF.IO/XGA6N to enable re-use of materials and improve reproducibility and transparency [@Marwick2017]. Also in this version-controlled compendium [@Marwick2018] are the raw data for all the visualizations and tests reported here and additional details of definition of our approach to tie formation, model fitting, MCMC run details, assessment of MCMC convergence and output, and goodness-of-fit diagnostics for our models. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.

@@

<!--posterior stats for pre-E-->
We examined the estimates from the posterior distributions to compare their differences in structure of the simulated networks (Table \@ref(tab:table-posterior-stats); Figure \@ref(fig:figure-posterior-distribution)). For nodal covariates in the pre-European model, the values of ritual element represented by pots (ritual-homophily) are mostly greater than zero. This suggests a positive effect on the formation of relations between burials, while the values of wealth rank represented by burial values (wealth-homophily) are mostly negative, indicating it has less influence on the tie formation. This demonstrates that burials with ritual pottery tend to form relations, but burials in the same wealth level tend not to. For other covariates with values mostly around zero, such as age (age-homophily) and sex (sex-homophily), we do not see any clear tendency. Most values are positive for the physical distance variable (dyadic covariate), indicating that physical proximity between burials influences similarity in burial treatment. For the endogenous network effects, transitivity presents a significant positive effect, while centralization demonstrates a negative effect. The high positive value for transitivity suggests a tendency of burials with similar burial goods to be clustered as connected communities, indicative of the presence of multiple corporate groups sharing burial goods in common. In contrast, the strong negative centralization shows there is a tendency toward decentralization that reflects most burials having a similar number of ties without any prominent burials. This might imply that individuals have equal access to trade goods in terms of the flow of goods.

@@

library(flextable)
# add column labels
ft <- flextable(posterior_esti_two_phases) %>% 
  colformat_double(digits = 2) %>% 
  set_header_labels(mean_pre = "pre-European", median_pre = "pre-European", 
                    `2.5%_pre` = "pre-European", `97.5%_pre` = "pre-European",
                    mean_post = "post-European", median_post = "post-European", 
                    `2.5%_post` = "post-European", `97.5%_post` = "post-European") %>%
  merge_at(i = 1, j = 2:5, part = "header") %>% 
  merge_at(i = 1, j = 6:9, part = "header") %>% 
  add_header_row(values = c("", "mean", "median", "2.5%", "97.5%",
                            "mean", "median", "2.5%", "97.5%"), top = FALSE ) %>% 
  align(j = 2:9, align = "right", part = "header") %>% 
  align(i = 1, align = "center", part = "header") %>% 
  width(j = ~ parameter, width = 1.36) %>% 
  width(j = 2:9, width = 0.67) %>% 
  set_caption("Estimated posterior means, medians, and 95% confidence intervals for each network parameter of two models. Confidence intervals that do not include zero indicate significant effects of parameters (Caimo, 2017). Posterior means close to posterior medians that indicate symmetric distributions.")
ft
```

@@

<!--posterior stats for for post-E and their comparison-->
For the post-European network model, two nodal covariates, wealth rank and age, both show tendencies towards positive values in the posterior distributions. This indicates the burials in the same wealth level and age groups tend to form relations. However, there is no clear pattern for ritual and sex variables since most values are around zero. There are also no effects for the variable of physical distance. Similar to the pre-European network, transitivity demonstrates a significant positive effect, but weaker than the effect of the pre-European network. In contrast, centralization has a significantly higher positive effect than the effect of the pre-European network. This means there is a tendency toward centralization, reflecting a limited number of burials having many more ties than others. This implies that the presence of better access to trade goods and the behavior of wealth accumulation and display in burial events. In general, the post-European network model has a smaller transitivity effect and higher centralization effect than the pre-European network model. This may suggest a reduced tendency toward clustering but high tendency toward centralization after the European presence.

@@

<!-- comparison between two networks using bootstrap --> 
One key difference between the pre-European and post-European networks is their size, with `r pre_E` burials compared to `r post_E` burials. To confirm the robustness of our comparison between these two different-sized networks, we used a vertex bootstrap technique to cross-validate the results of our Bayesian ERGMs. The vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. nodes) to quantify standard errors and estimate sampling variability in the network statistics of interest [@Chen2019; @Snijders1999; @Rroberts2021]. This enables the evaluation of uncertainty for networks and tests the difference between multiple networks of different sizes by examining their confidence intervals for the network population. We used the vertex bootstrap to compute endogenous network statistics, including density, centralization, and transitivity for our two networks. We explored the sample size effect by removing nodes at certain percentages (5-40%) for both networks and comparing their confidence intervals. The results demonstrate a consistent difference between the two networks up to 35% node removals for the centralization variable as shown in Figure \@ref(fig:figure-bootstrap-CI). This shows consistency with our earlier finding of negative centralization in the pre-European period and positive centralization in the post-European period, over several conditions of node removal. Similarly consistent with our earlier result, the vertex bootstrap shows no significant difference on density and transitivity. This indicates that the network variables, including density, centralization, and transitivity, are robust to the different network sizes of our sample.

@@

<!--archaeological implication based on the results-->
Network covariates give insights into the socio-cultural factors that were associated with the acquisition and distribution of foreign prestige goods. The general pattern of posterior distributions indicates that the main factor contributing to burial networks for the pre-European network was ritual, and for the post-European network was wealth. We interpret these results as illustrating different social and economic mechanisms influencing relations between burials for the two periods. Before European arrival, ritual behaviors could have been important factors that structured the interconnections of prestige goods in the burial contexts. This might imply that ritual was status indicator at the time before many foreign goods were introduced to Indigenous societies. After the European presence was established, status indicators shifted from ritual to wealth differences, corresponding with the characteristics of corporate/network strategies. In addition, geographical proximity of burials, implying kinship association, has some effect on sharing similar prestige items between burials in the pre-European period, but not in the post-European period. This may suggest that the use of prestige items is more individual-oriented instead of kinship-based after European arrival. Burials in the same age group tend to have similar goods after European arrival. However the sex of the buried individual did not have any effects on network formation for both networks. We should also note that their effects may be impacted by the unavailability of age (`r age_miss` out of `r pre_and_post` burials) and sex (`r sex_miss` out of `r pre_and_post` burials) values in our sample.

@@

<!-- trade dynamic during this period and future work for comparison -->
We recognize that complex local and international trade dynamics between different Indigenous groups and other foreign presence, Han Chinese and Japanese, might have contributed to social changes at Kiwulan as well. However, the current changes observed from burial networks and spatial distribution of trade items in the residential area at Kiwulan demonstrate consistent changes before and after the European arrival. The frequent trade activities stimulated by the presence of Europeans led to many trade items, including Anping jars, stonewares, Chinese porcelains, and beads found at archaeological sites with Dutch occupation at different parts of Taiwan, such as Fort Zeelandia. As a large village with continuous occupation since the 14th century in northeastern Taiwan, Kiwulan provides a special example to explore the reaction of Indigenous people living on the periphery to European colonial presence at a settlement scale. Was its experience with the European presence unique or typical of the region? We need more evidence from other sites in northeastern Taiwan to dive deeper into pericolonial effects, and compare evidence between sites on the periphery and colonial centers, such as Heping Dao and Tamsui. Future work should focus on collecting data from Indigenous sites related to European contact for comparison with Kiwulan.
